Estimate RSV onset and peak timing

### 2023-01-11 RSV_timing_explain/RSV_timing_tutorial.Rmd
add second derivatives

Relevant readings:
\
\
- [RSV onset timing at county level](https://doi.org/10.1093/cid/civ331)
\
\
- [RSV peak timing on state level](https://doi.org/10.1371/journal.ppat.1004591)
\
\
- [Comparing RSV onset timing before and during the COVID-19 pandemic](https://doi.org/10.1111/irv.12965)
\
\
- [RSV peak timing at ZIP code level and the drivers of RSV spread](https://doi.org/10.1126/sciadv.abd6421)

@@

>>>>

Relevant readings:
\
\
- [RSV onset timing at county level](https://doi.org/10.1093/cid/civ331)
\
\
- [RSV peak timing on state level](https://doi.org/10.1371/journal.ppat.1004591)
\
\
- [Comparing RSV onset timing before and during the COVID-19 pandemic](https://doi.org/10.1111/irv.12965)
\
\
- [RSV peak timing at ZIP code level and the drivers of RSV spread](https://doi.org/10.1126/sciadv.abd6421)
\
\
- [Assessment and optimization of respiratory syncytial virus prophylaxis in Connecticut, 1996–2013](https://www.nature.com/articles/s41598-021-90107-8)
\
\
- [Disease outbreak outcome estimation using penalized splines](https://cran.r-project.org/web/packages/pspline.inference/vignettes/seasonal.html)

@@

## Identify the onset of an epidemic using the second derivative method

When we look at the pseudo-RSV cases in the last 12 months, it is hard to determine when is the onset by eyes.

```{r, echo=F}
plot(1:12, tail(xt1,12),type="l", lwd=3, xaxt="n",bty="l",
      ylab="Cases of RSV", xlab="Month(n)", 
      ylim=range(c(xt1,xt2)),
     xlim=c(0,15))
lines(1:12,tail(xt2,12),col="red")
axis(2, at = seq(0,120,20))
axis(1, at = seq(0,12,1))
text(x=14, y=xt1[119], labels = "state1",col="black")
text(x=14, y=xt2[119], labels = "state2",col="red")
```

## Fit a regression to the observed RSV cases

Because now we only observe part of the epidemic, we cannot use the harmonic regression to fit what we observe. Instead, we use p-spline regression, a generalized additive model. Please check out [Smooth terms in GAM](https://stat.ethz.ch/R-manual/R-devel/library/mgcv/html/smooth.terms.html), [P-splines in GAMs](https://stat.ethz.ch/R-manual/R-devel/library/mgcv/html/smooth.construct.ps.smooth.spec.html) and Iris's work on [p-spline inference](https://cran.r-project.org/web/packages/pspline.inference/vignettes/seasonal.html) for introduction.

```{r}
require(mgcv)
irregular_model <- gam(cases ~ s(x=time, k=5, bs="ps"),
                       family=poisson, method="REML", data=data.frame(time=seq(1:12), cases=tail(xt1,12))) 
# k is knots
# The value for k set the upper limit on the 
# wiggliness of the smooth function
# You can choose the value based on AIC scores as well 
```

## Comparing the model fit and what we observed
The dots are what we observed and the line is the model fit
```{r,fig.dim=c(4, 4)}
plot(1:12, tail(xt1,12),type="p", xaxt="n",bty="l",
      ylab="Cases of RSV", xlab="Month(n)", 
      ylim=range(c(xt1,xt2)),
     xlim=c(0,12))
lines(1:12,fitted(irregular_model))
```

## Function to calculate the first and second derivatives
```{r}
deriv <- function(x, y) diff(y) / diff(x) # function to calculate derivative
middle_pts <- function(x) x[-1] - diff(x) / 2 # function to assist derivative calculation
```

```{r}
#install.packages("pspline.inference")
library(pspline.inference)
t <- seq(0.5,12,0.01)
dtime <- seq(0.5,12,0.01)

# we use Iris's package to get the uncertainty interval
# of disease trajectory. 
data=pspline.sample.timeseries(irregular_model, data.frame(time=t), pspline.outbreak.cases, samples=1)

deriv.pred = data.frame(deriv=diff(data$cases)/diff(data$time),time=c(1:length(diff(t)))) # calculate the first derivative
```
